Infant Birth Weight in Brazil

### 2024-08-21 code/analysis-code/fullanalysis-v1.qmd
Revisions for submission 4
Incorporated collinearity models

```{r, message=FALSE,  warning=FALSE}
#Path to summary data. Note the use of the here() package and not absolute paths
data_location <- here::here("data","processed-data","ML_linear.rds")
#load data
ML_linear <- readRDS(data_location)
```

@@

# Save regression output into table
saveRDS(ML_cv1_metrics, file = here::here("results", "tables", "ML_cv1_metrics.rds"))
```
Performing cross-validation results in an RMSE that is the same as our training model (488), lower than our full model. The r-squared (0.015) is also higher than our training and full models (0.01), demonstrating better model performance.

@@

# Merging Model 1 and 2 into Table

```{r}
# Combine the two tables using tbl_merge
table3 <- tbl_merge(
  list(table_model1, table_model2),
  tab_spanner = c("**Model 1**", "**Model 2**")) %>%
  modify_table_body(fun = ~.x %>% arrange(variable))
table3

>>>>

```{r, message=FALSE,  warning=FALSE}
#Path to summary data. Note the use of the here() package and not absolute paths
data_location <- here::here("data","processed-data","ML_linear.rds")
#load data
ML_linear <- readRDS(data_location)
str(ML_linear)
```

@@

# Save regression output into table
saveRDS(ML_cv1_metrics, file = here::here("results", "tables", "ML_cv1_metrics.rds"))
```
Performing cross-validation results in an RMSE that is the same as our training model (488), lower than our full model. The r-squared (0.017) is also higher than our training and full models (0.01), demonstrating better model performance.

@@

## Models 1 and 2 Sub-Samples

```{r}
# Model 1: Sub-sample simple model (ML_lm1_sub)
# Create a list to store R-squared values
r_squared_values_sub <- list()

# Filter data based on desired conditions
filtered_data <- ML_linear %>%
  filter(Brazilian == "Brazilian")

# Set reference level for "Euro-descent"
ML_linear$ModifiedColor <- relevel(ML_linear$ModifiedColor, ref = "Euro-Descent")

# Specify and fit model using linear_reg() (default engine, OLS) from tidymodels
ML_lm1_sub <- linear_reg() %>% set_engine("lm") %>% fit(Weightgrams ~ ModifiedColor, data = filtered_data)

# Summarize model fit with tidiers
ML_lm1_sub_summary <- summary(ML_lm1_sub) # Summary of fit
tidy(ML_lm1_sub, conf.int = TRUE) %>% mutate(p.value = round(p.value, 5)) # Coefficients and confidence intervals with p-values rounded to 5 decimal places
ML_lm1_sub_stat <- glance(ML_lm1_sub) # Extract model-level statistics

# Extract R-squared value
r_squared_values_sub$ML_lm1_sub <- ML_lm1_sub_stat$r.squared

# Create a gtsummary table directly from the model
ML_lm1_sub_table <- 
  tbl_regression(ML_lm1_sub, exponentiate = FALSE, intercept = TRUE) %>%
  modify_column_unhide(columns = c(statistic, std.error)) %>%
  add_significance_stars(hide_ci = TRUE, hide_p = TRUE) %>%
  add_glance_table(include = c(r.squared, nobs)) %>%
  modify_header(statistic = "**T-Statistic**") # rename header from statistic to t-statistic
ML_lm1_sub_table

# Save regression output into table
saveRDS(ML_lm1_sub_table, file = here::here("results", "tables", "ML_lm1_sub_table.rds"))
```
```{r}
# Model 2: Sub-sample multivariate model (ML_lm2_sub)
# Filter data based on desired conditions
filtered_data <- ML_linear %>%
  filter(Brazilian == "Brazilian")

# Set reference level for "Euro-descent"
ML_linear$ModifiedColor <- relevel(ML_linear$ModifiedColor, ref = "Euro-Descent")

# Specify and fit model using linear_reg() (default engine, OLS) from tidymodels
ML_lm2_sub <- linear_reg() %>% set_engine("lm") %>% fit(Weightgrams ~ ModifiedColor + Age + + ModifiedStatus, data = filtered_data)

# Summarize model fit with tidiers
ML_lm2_sub_summary <- summary(ML_lm2_sub) # Summary of fit
tidy(ML_lm2_sub, conf.int = TRUE) %>% mutate(p.value = round(p.value, 5)) # Coefficients and confidence intervals with p-values rounded to 5 decimal places
ML_lm2_sub_stat <- glance(ML_lm2_sub) # Extract model-level statistics

# Extract R-squared value
r_squared_values_sub$ML_lm2_sub <- ML_lm2_sub_stat$r.squared

# Create a gtsummary table directly from the model
ML_lm2_sub_table <- 
  tbl_regression(ML_lm2_sub, exponentiate = FALSE, intercept = TRUE) %>%
  modify_column_unhide(columns = c(statistic, std.error)) %>%
  add_significance_stars(hide_ci = TRUE, hide_p = TRUE) %>%
  add_glance_table(include = c(r.squared, nobs)) %>%
  modify_header(statistic = "**T-Statistic**") # rename header from statistic to t-statistic
ML_lm2_sub_table

# Save regression output into table
saveRDS(ML_lm2_sub_table, file = here::here("results", "tables", "ML_lm2_sub_table.rds"))
```

# Merging Model 1 and 2 and sub-sample 1 and 2 into Table

```{r}
# Combine the four tables using tbl_merge
table3 <- tbl_merge(
  list(table_model1, ML_lm1_sub_table, table_model2, ML_lm2_sub_table),
  tab_spanner = c("**Model 1**", "**Sub-Model 1**", "**Model 2**", "**Sub-Model 2**")) %>%
  modify_table_body(fun = ~.x %>% arrange(variable))
table3

### 2024-08-21 code/analysis-code/introanalysis-v1.qmd
Revisions for submission 4
Incorporated collinearity models

These results imply that Europeans had heavier infants than Brazilians. Since all women of color in this sample were Brazilian, there is a problem with multicollinearity. So I will perform some subsample modeling to understand this issue. We will create a `Brazilian` (yes, no) variable and then run a regression with this variable. Then, I will drop all non-Brazilians from the study and run the original regression with the covariates.

```{r}
# Create dummy variables for new category of Brazilian or not
# Create dummy variables for status and change to English
ML_linear$Brazilian <- recode_factor(ML_linear$ModifiedNationality, "European" = "Non-Brazilian", "Middle Eastern" = "Non-Brazilian", "Latin American" = "Non-Brazilian", "Brazilian" = "Brazilian")

str(ML_linear)
```

@@

# Model 6 sub-sample simple linear model
# Set reference level for Brazilian
ML_linear$ModifiedColor <- relevel(ML_linear$ModifiedColor, ref = "Euro-Descent")

>>>>

These results imply that Europeans had heavier infants than Brazilians. Since all women of color in this sample were Brazilian, there is a problem with multicollinearity. So I will perform some subsample modeling to understand this issue. We will create a `Brazilian` (yes, no) variable, drop all non-Brazilians from the study, and run the original regression with the covariates.

```{r}
# Create dummy variables for new category of Brazilian or not
# Create dummy variables for status and change to English
ML_linear <- mutate(ML_linear, 
                    Brazilian = recode_factor(ModifiedNationality, 
                                              "European" = "Non-Brazilian",
                                              "Middle Eastern" = "Non-Brazilian",
                                              "Latin American" = "Non-Brazilian",
                                              "Brazilian" = "Brazilian"))

# Check the structure of the updated data frame
str(ML_linear)
```

@@

# Model 6 sub-sample simple linear model
# Set reference level for "Euro-descent"
ML_linear$ModifiedColor <- relevel(ML_linear$ModifiedColor, ref = "Euro-Descent")

### 2024-08-21 products/manuscript/Manuscript_Final_Lancet.qmd
Revisions for submission 4
Incorporated collinearity models

```{r, echo=FALSE, message=FALSE}
#Loading packages
#install.packages("flextable")
#install.packages("gdtools")
library(tidyverse) #for data processing/cleaning
library(skimr) #for nice visualization of data 
library(here) #to set paths
library(renv) #for package management
library(gtsummary) #for summary tables
library(flextable) #for nice tables
library(gt) #for nice tables
```
# Abstract
## Background
In 1888 Brazil became the last country in the Western Hemisphere to abolish slavery after nearly four centuries of its existence. After abolition, the state did not create programs to incorporate people formerly enslaved and their descendants into economic, civil, or political life. Qualitative historical research has outlined the gendered and racialized health disparities of people of African descent at the turn-of-the-twentieth century. Public health studies have shown that health disparities continue to mirror patterns from over a century ago. The goal of this study is to quantify health disparities in infant birth weights by maternal characteristics in the post-abolition period in one of Brazil's first public maternity hospitals.

## Methods
This cross-sectional analysis uses hospital records on infants delivered between 1922-1926 at the Maternidade Laranjeiras (now Maternidade Escola) in Rio de Janeiro, Brazil, the city's first public maternity hospital. I used linear models to assess differences in infant birth weight by maternal skin color, age, number of pregnancies, and nationality.

## Findings
African ancestry was correlated with lower birth weights. Infants born to Afro-descendant women had birth weights estimated to be 85 grams lighter (p-value <0.01) than infants born to Euro-descendant women, whether Brazilian or immigrant. Among Afro-descendant women, infants born to Black (*preta*) women had birth weights estimated to be 100 grams lighter (p-value <0.01) than infants born to White women, and infants born to mixed-race (*parda*) women had birth weights estimated to be 72 grams lighter (p-value <0.05) than infants born to White women.

## Interpretations
The findings were likely the consequence of the historical legacies of slavery, particularly race-based socioeconomic inequality -- including more strenuous work schedules, poorer nutrition, and less sanitary living environments for women of color -- and possible epigenetic effects from the lived experience of racism. The findings are consistent with current-day research on racialized health disparities in Brazil, demonstrating the need to address the root causes of racism and its effects on health.

## Funding
This archival research for this project was funded by the National Science Foundation (NSF) and the Fulbright IIE Student Program.
{{< pagebreak >}}

@@

## Variables
### Outcome

*Birth weight*. The outcome was birth weight measured as a continuous variable in grams for live births only. In a secondary analysis available in the supplement, I also evaluated birth weight as a binary variable (either normal birth weight, NBW $>2500$ g or low birth weight, LBW $\leq2500$ g), according to current WHO classifications) [@worldhealthorganization2022]. The use of this cutoff in historical studies can be questionable as it was not even an established category in the 1920s. A Finnish pediatrician first proposed the 2500 g cutoff in 1919 without any specific justification. This was a period in which it was difficult for physicians to accurately calculate gestational age, and many physicians used birth weight as a proxy for prematurity. The WHO adopted this cutoff in 1948, and it has remained as the standard ever since [@hughes2017].

@@

Except at the extremes, racial classification in Brazil was, and continues to be, imprecise as racial categories and skin color classifications in Brazil exist on a spectrum rather than as a Black-White binary as in the United States [@travassos2004]. Using more general categories addresses imprecision by grouping all possible mothers of any African descent into one category. This method is in line with recent studies, which look at all African-descended peoples as a group and then stratify by racial mixing [@nyarko2013; @wehby2015a].

@@

I calculated summary statistics for the sample, including sample proportions for maternal skin color, maternal ancestry, parity, maternal nationality, birth outcome, maternal outcome, fetal outcome, and fetal sex. I also calculated mean maternal age, infant birth length, and infant birth weight, as a whole and stratified by fetal sex and maternal skin color.

@@

I used Ordinary Least Squares (OLS) to establish a relationship of maternal variables and infant birth weight. I employ birth weight as a continuous variable in line with both current and historical studies [@andrasfay2020a; @costa1998; @costa2004; @david1997; @leal2006; @silveira2019]. Due to the lack of comprehensive information on gestational age in these data, I did not consider birth weight in reference to gestational age, as is best practice for disentangling PTB from LBW [@cutland2017]. I discuss this constraint in the discussion below.

@@

@@

The overall rate of LBW for infants born to Afro-descended mothers was 12.4%, compared to 9.7% for Euro-descended mothers. A difference in means test showed borderline significance (t-statistic = 1.9119, df. 2008.2, p-value 0.056, [95% CI -0.0007, 0.0532]).

@@

In models 1 and 2 (Table 2), women who were nulliparous (delivering their first child) had infants weighing on average 113 grams less than multiparous women. In models 3 and 4 (Table 3), the average was 114 grams lighter. This is an association found both in other historical studies and today [@costa1998; @souza2020]. Age and nationality were not significant.

@@

@@

All models based on historical data must contend with limited sample sizes, missing data, and imprecise measures of health and nutrition [@franken2019; @steckel1998; @roberts2014]. In this study, a major limitation was the original published clinical notes' exclusion of gestational age. Thus, the birth weight variable captured both smaller babies born at term (low weight at-term infants, or small for gestational age, SGA) and pre-term infants, or those born before 37 weeks' gestation. We are thus unable to know how many of these infants were low birth weight at term [@falcao2020]. This limits our ability to understand the relationship between maternal variables and infant birth weight as pre-term infants are more likely to have lower birth weights [@cutland2017]. Given the absence of accurate technological techniques for determining gestational age in the past, this limitation is inherent to any historical inquiry into birth weights, although other historical studies have datasets that include gestational age [@costa1998; @costa2004; @goldin1989]. I assume that infants at risk for PTB were also at-risk for LBW. WE CANNOT MEASURE THE VARIABLE THAT IS REALLY DRIVING THINGS.

Lack of data on maternal nutrition and health status, which cannot be corrected for by using maternal skin color as a proxy for overall health, is a major impediment to understanding the true relationship between maternal variables and infant birth weight in this study. Given the paucity of historical understandings of current-day health disparities in maternal-infant health, this paper demonstrates the need for more research into quantitative associations between socio-demographic categories and health in the past.

>>>>

```{r, echo=FALSE, message=FALSE}
#Loading packages
#install.packages("flextable")
#install.packages("gdtools")
library(tidyverse) #for data processing/cleaning
library(skimr) #for nice visualization of data 
library(here) #to set paths
library(renv) #for package management
library(gtsummary) #for summary tables
library(flextable) #for nice tables
library(gt) #for nice tables
```

# Abstract

## Background

In 1888 Brazil became the last country in the Western Hemisphere to abolish slavery after nearly four centuries of its existence. After abolition, the state did not create programs to incorporate people formerly enslaved and their descendants into economic, civil, or political life. Qualitative historical research has outlined the gendered and racialized health disparities of people of African descent at the turn-of-the-twentieth century. Public health studies have shown that health disparities continue to mirror patterns from over a century ago. The goal of this study is to quantify health disparities in infant birth weights by maternal characteristics in the post-abolition period in one of Brazil's first public maternity hospitals.

## Methods

This cross-sectional analysis used hospital records on infants delivered between 1922-1926 at the Maternidade Laranjeiras (now Maternidade Escola) in Rio de Janeiro, Brazil, the city's first public maternity hospital. I used linear models to assess differences in infant birth weight by maternal skin color, age, number of pregnancies, and nationality.

## Findings

African ancestry was correlated with lower birth weights. Infants born to Afro-descendant women had birth weights estimated to be 85 grams lighter (p-value <0.01) than infants born to Euro-descendant women, whether Brazilian or immigrant. Among Afro-descendant women, infants born to Black (*preta*) women had birth weights estimated to be 100 grams lighter (p-value <0.01) than infants born to White women, and infants born to mixed-race (*parda*) women had birth weights estimated to be 72 grams lighter (p-value <0.05) than infants born to White women.

## Interpretations

The findings were likely the consequence of the historical legacies of slavery, particularly race-based socioeconomic inequality -- including more strenuous work schedules, poorer nutrition, and less sanitary living environments for women of color -- and possible epigenetic effects from the lived experience of racism. The findings are consistent with current-day research on racialized health disparities in Brazil, demonstrating the need to address the root causes of racism and its effects on health.

## Funding

This archival research for this project was funded by the National Science Foundation (NSF) and the Fulbright IIE Student Program.
{{< pagebreak >}}

@@

## Variables

### Outcome

*Birth weight*. The outcome was birth weight measured as a continuous variable in grams for live births only. In a secondary analysis available in the supplement, I also evaluated birth weight as a binary variable (either normal birth weight, NBW $>2500$ g or low birth weight, LBW $\leq2500$ g), according to current WHO classifications) [@worldhealthorganization2022]. We can question the use of this cutoff in historical studies, as it was not an established category until the mid-twentieth century. A Finnish pediatrician first proposed the 2500 g cutoff in 1919 without any specific justification. This was a period in which it was difficult for physicians to accurately calculate gestational age, and thus many physicians used birth weight as a proxy for prematurity [@wilcox CITE CORRECTLY]. The WHO adopted this cutoff in 1948, and it has remained the standard since [@hughes2017].

@@

Except at the extremes, racial classification in Brazil was, and continues to be, imprecise, as racial categories and skin color exist on a spectrum rather than as a Black-White binary as in the United States [@travassos2004]. Using more general categories addresses imprecision by grouping all possible mothers of any African descent into one category. This method is in line with recent studies, which look at all African-descended peoples as a group and then stratify by racial mixing [@nyarko2013; @wehby2015a].

@@

I calculated summary statistics, including sample proportions for maternal skin color, maternal ancestry, parity, maternal nationality, birth outcome, maternal outcome, fetal outcome, and fetal sex. I also calculated mean maternal age, infant birth length, and infant birth weight, unadjusted and then stratified by first fetal sex and then maternal skin color.

@@

I used Ordinary Least Squares (OLS) to establish a relationship of maternal variables and infant birth weight. I employ birth weight as a continuous variable in line with both current and historical studies [@andrasfay2020a; @costa1998; @costa2004; @david1997; @leal2006; @silveira2019]. Due to the lack of comprehensive information on gestational age in these data, I did not consider birth weight in reference to gestational age. I discuss this constraint in the discussion below.

@@

Scholars have shown that immigrants to Brazil during this historical period were healthier and better off socio-economically than their Brazilian counterparts, both of color and White [@franken2019; @ball2020]. Given the possibility for collinearity between maternal nationality and maternal skin color, I ran a sub-sample analysis for Models 1 and 2 in two steps. First, I created a new variable `Brazilian` (yes/no) and ran a simple linear analysis with birth weight as the dependent variable and maternal nationality as the independent variable. Non-Brazilian women birthed infants that weighed on average 100 g more than Brazilian women (the reference group) (p-value <0.001). Given that all women of color in the data set were Brazilian, there is collinearity occurring. Thus, second, I dropped all non-Brazilians from the study, and ran the original regressions from Models 1 and 2.

@@

The overall rate of LBW for infants born to Afro-descended mothers was 12.4%, compared to 9.7% for Euro-descended mothers. A difference in means test showed borderline significance (t-statistic = 1.9119, df. 2008.2, p-value 0.056, [95% CI -0.0007, 0.0532]). Regardless of a difference between infants born to mothers of different skin colors, all LBW infants in the past, the probability of surviving the neonatal period was lower than today given higher neonatal and infant mortality rates [@costa1998; @roth2020].

@@

UPDATE FOR CORRELATION STUFF

In models 1 and 2 (Table 2), women who were nulliparous (delivering their first child) had infants weighing on average 113 grams less than multiparous women. In models 3 and 4 (Table 3), the average was 114 grams lighter. This is an association found both in other historical studies and today [@costa1998; @souza2020]. Age and nationality were not significant. UPDATE

@@

WHO: birth weight "is strongly conditioned by the health and nutritional status of the mother"

WARD: "In addition, because birth weight is sensitive to environmental factors during the nine months of gestation (especially the final three months) as well as the long-range effects of changes in living conditions, it reflects both short- and long-term trends." (5)

WARD: "By and large, the most privileged social groups bear the heaviest babies while the most disadvantaged have the lightest. These differences, in turn, reflect the fact that fetal growth and weight at birth are influenced by a wide range of biological, social, and economic factors, not all of which are fully understood." (10)

WARD: From the standpoint of neonatal mortality the optimum weight of a newborn child lies between 3500 and 3900 grams. Infants bom outside this range have higher mortality levels than do those within it. The riSks are especially great among those weighing considerably less than the optimum. Estimates suggest that the chance of death during the first week of life increases by 40 percent, and the risk of fetal death by 25 percent, for every 100-gram decrease below the optimum level of weight. (10)

Optimal birth weight is the weight at which birth weight-specific mortality is lowest

@@

All models based on historical data must contend with limited sample sizes, missing data, and imprecise measures of health and nutrition [@franken2019; @steckel1998; @roberts2014; @ward1993]. In this study, a major limitation was the original published clinical notes' exclusion of gestational age. Thus, the birth weight variable captured both smaller babies born at term (low weight at-term infants, or small for gestational age, SGA) and pre-term infants, or those born before 37 weeks' gestation. We are thus unable to know how many of these infants were low birth weight at term [@falcao2020]. This limits our ability to understand the relationship between maternal variables and infant birth weight as pre-term infants are more likely to have lower birth weights [@cutland2017]. Given the absence of accurate technological techniques for determining gestational age in the past, this limitation is inherent to any historical inquiry into birth weights, although other historical studies have datasets that include gestational age [@costa1998; @costa2004; @goldin1989]. I assume that infants at risk for PTB were also at-risk for LBW.

Lack of data on maternal nutrition and health status or maternal work routine and income, all of which cannot be corrected for by using maternal skin color as a proxy, is a major impediment to understanding the true relationship between maternal variables and infant birth weight in this study. Maternal skin color may serve as a partial index of the unobservable variables that are acting upon infant birth weight [@costa2004; @ward1993]. Given the paucity of historical understandings of current-day health disparities in maternal-infant health, this paper demonstrates the need for more research into quantitative associations between socio-demographic categories and health in the past.

HOW INFANTS WERE WEIGHED AND ACCURACY OF SCALE
